---
title: "Reanalysis of Vernier et al. 2020, Science Advances"
author: "Joanito Liberti, University of Lausanne"
date: 
output: github_document
---

# We analyze data from the live vs. heat-killed inoculum experiment in Vernier et al. 2020 Science Advances
```{r setup, include = FALSE}
# Set the working directory (in R notebook the working directory needs to be assigned within a "setup" chunk, or it will only work within the individual chunks in which it was set)

path <- "/Volumes/gr_Engel/lab_resources/Manuscripts/2023_Liberti_Maturation/Code/CHC_profiles/"
knitr::opts_knit$set(root.dir = normalizePath(path)) 
```

# Load packages
```{r}
library(DESeq2)
library(ggplot2)
library(vegan) # ecological diversity analysis
library(dplyr)
library(scales) # scale functions for visualizations
library(grid)
library(RColorBrewer)
library(reshape2) # data manipulation package
library(cowplot)
library(phyloseq)
library(tidyverse)
library(readxl)
library(genefilter)
```

# Heatmap of CHC profiles
```{r, echo=FALSE, message=FALSE, warning=FALSE}
newbee<-read.csv("CHCs_HeatkilledVsLiveExp.csv",row.names = 1) # data underlying Fig. 2E in Vernier et al. 2020 Science Advances - can be found in supplementary file abd3431_data_file_s1.xlsx (https://www.science.org/doi/10.1126/sciadv.abd3431#supplementary-materials)
bee2<-newbee[,-1]
bee2 <-bee2[1:(length(bee2)-1)] # Removing last column because it is the row sum

bee2 <- bee2/rowSums(bee2) # Transform into proportions

library(vegan)
library(ggplot2)

# Microbiome treatment
microbio<-newbee[,1]

## Heatmap
### calculate clustering method
library(dendextend)
library(colorspace)
dbee<-dist(bee2,method="euclidean")
hbee<-hclust(dbee,method="ward")

dend<-as.dendrogram(hbee)

dendSampleOrder <- labels(dend) # Save sample names in the same order as the dendrogram for later

gg_color_hue <- function(n) {
  hues = seq(15, 375, length = n + 1)
  hcl(h = hues, l = 65, c = 100)[1:n]
}

n = 2
cols = gg_color_hue(n)

microbio2<-as.factor(microbio)
labels_colors(dend) <- cols[
    as.numeric(microbio2)[order.dendrogram(dend)]
  ]

labels(dend) <- paste(as.character(microbio)[order.dendrogram(dend)],
                      "(",labels(dend),")", 
                      sep = "")
dend <- hang.dendrogram(dend,hang_height=0.1)

dend <- dendextend::set(dend, "labels_cex", 0.5)
dend %>% dendextend::set("branches_lwd", 1)

par(mar = c(3,3,3,7))
{plot(dend,  
     main = "Honeybee CHC-Microbiota: Euclidean distances with Ward's Hierarchical clustering", 
     horiz =  TRUE,  nodePar = list(cex = .007))

microbio_levels <- c("Heat killed", "Live")
legend("topleft", legend = microbio_levels, fill = cols)}

# Plot heatmap
library(gplots)
library(RColorBrewer)

nb.cols <- 100
mycolors <- colorRampPalette(brewer.pal(9, "Oranges"))(nb.cols)

microlab<-cols[as.factor(newbee$Treatment)]

dev.new()
pdf(file="CHCheatmap-LiveVsHeatKilled.pdf",width=4,height=4, pointsize=8)
{gplots::heatmap.2(as.matrix(bee2), 
                  main = "Relative percentages clustering of microbiome CHC Honeybees",
                  srtCol = 90,
                  dendrogram = "row",
                  Rowv = dend,
                  Colv = "NA", # this to make sure the columns are not ordered
                  trace="none",          
                  margins =c(5,0.1),      
                  key.xlab = "Relative Percentage on Cuticle",
                  RowSideColors = microlab, # to add nice colored strips        
                  col = mycolors)
legend("topright", legend = microbio_levels, fill = cols)}
dev.off()

{gplots::heatmap.2(as.matrix(bee2), 
                  main = "Relative percentages clustering of microbiome CHC Honeybees",
                  srtCol = 90,
                  dendrogram = "row",
                  Rowv = dend,
                  Colv = "NA", # this to make sure the columns are not ordered
                  trace="none",          
                  margins =c(5,0.1),      
                  key.xlab = "Relative Percentage on Cuticle",
                  RowSideColors = microlab, # to add nice colored strips        
                  col = mycolors)
legend("topright", legend = microbio_levels, fill = cols)}
```

## The heatmap reveals bees with two physiologies: nurses and foragers. We can see in the heatmap plot that the distribution of these bee types is unbalanced between gut microbiota treatments. However, the relative contribution of the two factors to the CHC profiles cannot be teased apart because cage replication is insufficient (2 per treatment in this experiment, A and B in sample names).

# NMDS plots
```{r, echo=FALSE, message=FALSE, warning=FALSE}
# We derive nurses and foragers based on the CHC clusters in the heatmap
NandF<-c(rep("nurse", each=15),rep("forager", each=17))
NandF<-data.frame(dendSampleOrder,NandF)
rownames(NandF)<-NandF[,1]

beetr2<-data.frame(bee2,microbio)
beetr2$Sample_ID <- rownames(beetr2)

beeTest <- merge(beetr2, NandF, by=0, all=TRUE)
beeTest[,c("Sample_ID","dendSampleOrder")]<-NULL
rownames(beeTest)<-beeTest[,1]
beeTest[,1]<-NULL
beeTest$Cage <- c(rep("Cage1", 8), rep("Cage2", 8),rep("Cage3", 8), rep("Cage4", 8) )

set.seed(68827)
mds1 <- metaMDS(as.matrix(beeTest[1:(length(beeTest)-3)]), distance="bray", k=2, trymax=100, autotransform=TRUE, noshare=0.1, expand=TRUE, trace=1)

## create table for NMDS X and Y values of chemical compounds
data.scores<-as.data.frame(scores(mds1)$sites)
data.scores$site<-rownames(data.scores)

## create table for NMDS X and Y values of individuals
species.scores<-as.data.frame(scores(mds1,"species"))
species.scores$species<-rownames(species.scores)

gg_color_hue <- function(n) {
  hues = seq(15, 375, length = n + 1)
  hcl(h = hues, l = 65, c = 100)[1:n]
}

n = 2
cols = gg_color_hue(n)

#Graph
p1 <- ggplot(col = cols) + 
  geom_point(data=data.scores,aes(x=NMDS1,y=NMDS2,color=beeTest$microbio),size=2) +
  coord_equal() +
  theme_bw() +
  theme(axis.title.x = element_text(size=10), # remove x-axis labels
        axis.title.y = element_text(size=10), # remove y-axis labels
        panel.background = element_blank(), 
        panel.grid.major = element_blank(),  #remove major-grid labels
        panel.grid.minor = element_blank(),  #remove minor-grid labels
        plot.background = element_blank())

p2 <- ggplot(col = cols) + 
  geom_point(data=data.scores,aes(x=NMDS1,y=NMDS2,color=beeTest$NandF),size=2) +
  coord_equal() +
  theme_bw() +
  color_palette(c("#56B4E9","#E69F00")) +
  theme(axis.title.x = element_text(size=10), # remove x-axis labels
        axis.title.y = element_text(size=10), # remove y-axis labels
        panel.background = element_blank(), 
        panel.grid.major = element_blank(),  #remove major-grid labels
        panel.grid.minor = element_blank(),  #remove minor-grid labels
        plot.background = element_blank())

library(ggpubr)
ggarrange(p1, p2, ncol=1)
ggsave(height=6,width=5,dpi=300, filename="HeatKilledVsLive_ordinations.pdf", useDingbats=FALSE)
```

# Microbiota stacked barplots
```{r, echo=FALSE, message=FALSE, warning=FALSE}
df = read.table("SciAdv-Microbiota-heatkilled.txt", header = T, sep="\t") # data underlying Fig. 2D in Vernier et al. 2020 Science Advances - can be found in supplementary file abd3431_data_file_s2.xlsx

df[, -c(1,2)] <- df[ , -c(1,2)]/1794 # Here we divide by 1794 to get proportions as the authors deposited the tables as supplement after rarefying to the same depth

# Here we filter out ASVs that have less than 2% relative abundance in two samples, to make barplots more legible
f1 <- kOverA(2, 0.02)
ffun <- filterfun(f1)
filt1 <- genefilter(t(df[, -c(1,2)]), ffun)

mdf<-df[, -c(1,2)][,filt1]

mdf <- cbind(Sample_ID = df$Sample_ID, Treatment = factor(df$Treatment), mdf)


mdf<- reshape(mdf, 
        direction = "long",
        varying = list(names(mdf)[3:ncol(mdf)]),
        v.names = "Value",
        idvar = c("Sample_ID", "Treatment"),
        timevar = "ASV", 
        times = colnames(mdf[3:ncol(mdf)]))

mdf$ASV<-sub("*Firm.5", "Firm 5", mdf$ASV)
mdf$ASV<-sub("*Firm.4", "Firm 4", mdf$ASV)
mdf$ASV<-sub("*\\.[0-9]", "", mdf$ASV)

library(RColorBrewer)
cbPalette<-colorRampPalette(brewer.pal(12, "Paired"))

set.seed(2) # Change the number within parentheses to change the order of colors in the plot. Here we use set.seed to make the colors reproducible, because in scale_fill_manual we use "sample" to shuffle the color order

mdf$Treatment <- factor(mdf$Treatment)
mdf$ASV <- factor(mdf$ASV)

p3 = ggplot(mdf, aes_string(x = "Sample_ID", y = "Value", fill = "ASV")) +
  geom_bar(stat = "identity", position = "stack", color = "black") +
  ylab("Relative abundance") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5), legend.position="bottom", strip.background = element_blank()) + # ,  strip.text.x = element_blank()
  facet_grid(. ~ Treatment, scale="free", space="fixed") +
  scale_fill_manual(values=sample(cbPalette(16)))  

print(p3, width = 1000, height = 200)

dev.new()
pdf(file="Barplots_LiveVsHeatKilled.pdf",width=10,height=4, pointsize=4)
p3
dev.off()
```

# We analyze what ASVs are signficantly different between treatment groups by DESeq2
```{r, echo=FALSE, message=FALSE, warning=FALSE}
df = read.table("SciAdv-Microbiota-heatkilled.txt", header = T, sep="\t") 

# Set plotting theme
theme_set(theme_bw())

#Create a phyloseq object
otutab<-df[,-2]
rownames(otutab)<-otutab[,1]
otutab[,1]<-NULL

mapping<-df[,c(1,2)]
rownames(mapping)<-mapping[,1]

taxa<- colnames(otutab)
taxa<-as.data.frame(taxa)
rownames(taxa) <- taxa[,1]
taxa<-as.matrix(taxa)

ps <- phyloseq(otu_table(otutab, taxa_are_rows=FALSE), 
               sample_data(mapping), 
               tax_table(taxa))

# Load DESeq2
library("DESeq2"); packageVersion("DESeq2")

diagdds = phyloseq_to_deseq2(ps, ~Treatment)

# calculate geometric means prior to estimate size factors
gm_mean = function(x, na.rm=TRUE){
  exp(sum(log(x[x > 0]), na.rm=na.rm) / length(x))
}
geoMeans = apply(counts(diagdds), 1, gm_mean)
diagdds = estimateSizeFactors(diagdds, geoMeans = geoMeans)
diagdds = DESeq(diagdds, fitType="local")

res = results(diagdds, cooksCutoff = FALSE)
alpha = 0.05
sigtab = res[which(res$padj < alpha), ]
sigtab = cbind(as(sigtab, "data.frame"), as(tax_table(ps)[rownames(sigtab), ], "matrix"))
sigtab

write.table(sigtab, "LiveVsHeatKilled-DESeq2.txt", sep="\t")
```

## Only 5 ASVs mostly representing opportunistic infections have significantly different relative abundances between heat-killed and live inoculation treatments in this experiment. 

# **Mono-inoculations experiment** - Microbiota Fig. 3B
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=if (fig_small) width_small else width_large}
df = read.table("SciAdv_Microbiota-3B-Monoinoculations.txt", header = T, sep="\t") # data underlying Fig. 3B in Vernier et al. 2020 Science Advances - can be found in supplementary file sciadv.abd3431_data_file_s2

df[, -c(1,2)] <- df[ , -c(1,2)]/rowSums(df[1, -c(1,2)]) # Here we divide by the row sum to get proportions as the authors deposited the table as supplement after rarefying to the same depth

# Here we filter out ASVs that have less than 2% relative abundance in two samples, to make barplots more legible
f1 <- kOverA(2, 0.02)
ffun <- filterfun(f1)
filt1 <- genefilter(t(df[, -c(1,2)]), ffun)

mdf<-df[, -c(1,2)][,filt1]

mdf <- cbind(Sample_ID = df$Sample_ID, Treatment = factor(df$Treatment), mdf)


mdf<- reshape(mdf, 
        direction = "long",
        varying = list(names(mdf)[3:ncol(mdf)]),
        v.names = "Value",
        idvar = c("Sample_ID", "Treatment"),
        timevar = "ASV", 
        times = colnames(mdf[3:ncol(mdf)]))

library(RColorBrewer)
cbPalette<-colorRampPalette(brewer.pal(12, "Paired"))

set.seed(2) # Change the number within parentheses to change the order of colors in the plot. Here we use set.seed to make the colors reproducible, because in scale_fill_manual we use "sample" to shuffle the color order.

mdf$Colony <- factor(mdf$Treatment)
mdf$ASV <- factor(mdf$ASV)

p3 = ggplot(mdf, aes_string(x = "Sample_ID", y = "Value", fill = "ASV")) +
  geom_bar(stat = "identity", position = "stack", color = "black") +
  ylab("Relative abundance") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5), legend.position="bottom", strip.background = element_blank()) + # ,  strip.text.x = element_blank()
  facet_grid(. ~ Colony, scale="free", space="fixed") +
  scale_fill_manual(values=sample(cbPalette(20)))  

print(p3, width = 1000, height = 200)

dev.new()
pdf(file="Barplots_Mono-inoculations.pdf",width=10,height=4, pointsize=4)
p3
dev.off()
```

# Let's analyze what ASVs are signficantly different between monoinoculations by DESeq2
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=if (fig_small) width_small else width_large}
df = read.table("SciAdv_Microbiota-3B-Monoinoculations.txt", header = T, sep="\t") 

# **Load and filter the data in phyloseq**
library(ggplot2)
library(vegan) # ecological diversity analysis
library(dplyr)
library(scales) # scale functions for vizualizations
library(grid)
library(RColorBrewer)
library(reshape2) # data manipulation package
library(cowplot)
library(phyloseq)
library(tidyverse)
library(readxl)

# Set plotting theme
theme_set(theme_bw())

#Create a phyloseq object
otutab<-df[,-2]
rownames(otutab)<-otutab[,1]
otutab[,1]<-NULL

mapping<-df[,c(1,2)]
rownames(mapping)<-mapping[,1]

taxa<- colnames(otutab)
taxa<-as.data.frame(taxa)
rownames(taxa) <- taxa[,1]
taxa<-as.matrix(taxa)

ps <- phyloseq(otu_table(otutab, taxa_are_rows=FALSE), 
               sample_data(mapping), 
               tax_table(taxa))
               
# sample_data(ps) 

# Load DESeq2
library("DESeq2"); packageVersion("DESeq2")

diagdds = phyloseq_to_deseq2(ps, ~ Treatment)

# calculate geometric means prior to estimate size factors
gm_mean = function(x, na.rm=TRUE){
  exp(sum(log(x[x > 0]), na.rm=na.rm) / length(x))
}
geoMeans = apply(counts(diagdds), 1, gm_mean)
diagdds = estimateSizeFactors(diagdds, geoMeans = geoMeans)
diagdds = DESeq(diagdds, fitType="local")

res = results(diagdds, cooksCutoff = FALSE)
alpha = 0.05
sigtab = res[which(res$padj < alpha), ]
sigtab = cbind(as(sigtab, "data.frame"), as(tax_table(ps)[rownames(sigtab), ], "matrix"))
sigtab
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# save.image (file = "SciAdvReanalysis.RData")
sessionInfo()
```


